---
title: My dynamic report
output: html_fragment
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, error=FALSE)
library(tidyverse)
```

```{css}
@import url('https://fonts.googleapis.com/css2?family=DM+Serif+Text:ital@0;1&family=Playfair+Display:ital,wght@0,400..900;1,400..900&display=swap');

@import url('https://fonts.googleapis.com/css2?family=DM+Serif+Text:ital@0;1&family=Lexend:wght@100..900&family=Playfair+Display:ital,wght@0,400..900;1,400..900&display=swap');

h2 {
  background-color: #f8766d;
  color: #ebebeb;
  padding: 20px;
}

h1, h3 {
  color: #f8766d;
  padding: 20px;
}

h1, h2, h3 {
  font-family: 'Playfair Display';
  font-weight: bold;
  text-align: left;
}

p, ul, ol {
  font-family: 'Lexend', sans-serif;
  line-height: 1.6;
  border-style: solid;
  border-color: #00bfc4;
  border-width: medium;
  padding: 20px;
}

body {
  background-color: #ebebeb;
}
```

## Commute Data Analysis

This report analyses data regarding my commute to/from the University of Auckland 
over the last few weeks. Data on which mode of transport/bus route I took, the 
crowded-ness of the bus, and the start time and duration of my commute was 
collected via a Google Form. This data was analysed using the R techniques taught 
in the Stats 220 course to determine trends and possibly inform suggestions on 
when/how I should commute to and from the University in the future.

```{r}
csv_file <- "https://docs.google.com/spreadsheets/d/e/2PACX-1vTRZ24GyKC8PL-ZaSI4BG6Y9DQVwao8TolkFlGLQr2Fc04ZeyEH2wudkVSBNBde7JGYko3bP3auaHTi/pub?output=csv"

logged_data <- read_csv(csv_file) 


########## PART D: Exploring Data in R ##########
# Rename columns:
latest_data <- logged_data %>%
  rename(
    "direction" = 3, 
    "transport_mode" = 4, 
    "start_time" = 5, 
    "duration" = 6, 
    "bus_route" = 7, 
    "crowded" = 8
    ) %>%
  mutate(start_time_seconds = as.numeric(start_time))

####### Summary values:
## Helper function to format time in seconds (i.e. 30840) into hh:mm:ss (i.e. 08:34:00)
num_to_time <- function(time_in_seconds) {
  hrs <- sprintf("%02d", time_in_seconds %/% 3600)
  mins <- sprintf("%02d", (time_in_seconds %% 3600) %/% 60)
  secs <- sprintf("%02d", time_in_seconds %% 60)
  return(paste0(hrs, ":", mins, ":", secs))
}

## Earliest, latest, shortest, longest trips to/from UoA home
earliest <- num_to_time(min(latest_data$start_time_seconds))
latest <- num_to_time(max(latest_data$start_time_seconds))
shortest <- min(latest_data$duration)
longest <- max(latest_data$duration)

## Average commute duration by mode of transport
avg_duration_by_mode <- latest_data %>%
  group_by(transport_mode) %>%
  summarise(avg_duration = mean(duration, na.rm = TRUE), .groups = "drop")

## Average commute duration by route and direction
avg_duration_by_route <- latest_data %>%
  mutate(bus_route = ifelse(is.na(bus_route), "Car", bus_route)) %>%
  group_by(bus_route, direction) %>%
  summarise(avg_duration = mean(duration), .groups = "drop", 
            avg_start_time = num_to_time(mean(start_time_seconds))) %>%
  arrange(direction, desc(avg_duration))
### Longest bus route overall (on average)
longest_overall <- avg_duration_by_route %>%
  filter(bus_route != "Car") %>%
  filter(avg_duration == max(avg_duration)) 
### Shortest bus route overall (on average)
shortest_overall <- avg_duration_by_route %>%
  filter(bus_route != "Car") %>%
  filter(avg_duration == min(avg_duration)) 

## Crowded-ness and average start time of each bus route based on the direction of travel
busiest_route <- latest_data %>%
  filter(!is.na(bus_route)) %>%
  group_by(bus_route, direction) %>%
  summarise(avg_crowded = mean(crowded, na.rm = TRUE), .groups = "drop",
            avg_start_time = num_to_time(mean(start_time_seconds))) %>%
  arrange(direction, desc(avg_crowded))
### Least crowded by direction
least_crowded_overall <- busiest_route %>%
  group_by(direction) %>%
  filter(avg_crowded == min(avg_crowded))
### Most crowded by direction
most_crowded_overall <- busiest_route %>%
  group_by(direction) %>%
  filter(avg_crowded == max(avg_crowded))

```


The earliest trip made during the data collection stage was at `r earliest`, 
and the latest was at  `r latest`. The shortest trip took `r shortest` minutes, 
while the longest took `r longest` minutes.

The collected data also shows that the `r avg_duration_by_mode$transport_mode[1]` 
is faster on average than the `r avg_duration_by_mode$transport_mode[2]` by 
`r abs(round(avg_duration_by_mode$avg_duration[2] - avg_duration_by_mode$avg_duration[1], 2))` 
minutes for my personal commute. 

This is interesting to note, however this figure alone does not provide much insight 
into my commute. Without taking into account factors like the time of my commute 
or the crowded-ness of the bus, I cannot get a good understanding of the trends 
that could inform which route I should take. Hence, we will now consider some of 
these factors:

```{r}
## Avg crowdedness by direction for each bus route
ggplot(busiest_route, aes(x = bus_route, y = avg_crowded, fill = direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Average Crowded-ness by Bus Route and Direction",
    x = "Bus Route",
    y = "Average Crowded-ness (0-5)",
    fill = "Direction"
  )
```


As shown above, the trip that is the least crowded commute from 
`r least_crowded_overall$direction[1]` tends to be the `r least_crowded_overall$bus_route[1]` bus 
route, with an average crowded-ness rating of `r round(least_crowded_overall$avg_crowded[1], 2)`/5 
(where 1 = Empty Bus, and 5 = Full Bus). The average start time of the recorded 
data from this trip is `r (busiest_route %>% filter(bus_route == least_crowded_overall$bus_route[1], direction == least_crowded_overall$direction[1]))$avg_start_time`.

The trip that is the least crowded commute from 
`r least_crowded_overall$direction[2]` is, on average, the `r least_crowded_overall$bus_route[2]` bus 
route, with an average crowded-ness rating of `r round(least_crowded_overall$avg_crowded[2], 2)`/5. 
The average start time of the recorded 
data from this trip is `r (busiest_route %>% filter(bus_route == least_crowded_overall$bus_route[2], direction == least_crowded_overall$direction[2]))$avg_start_time`.

Thus, in future I should consider taking these routes in order to better my chances 
of getting a seat and commuting comfortably.

However, it is also useful to consider the efficiency of these different bus routes:

```{r}
## Avg commute duration by bus route and direction
ggplot(avg_duration_by_route, aes(x = bus_route, y = avg_duration, fill = direction)) +
  stat_summary(geom = "bar", position = "dodge") +
  labs(
    title = "Average Travel Duration by Route and Direction",
    x = "Route",
    y = "Average Duration (mins)",
    fill = "Direction"
  ) 

```

From the plot above, we can see that the quickest bus route from `r least_crowded_overall$direction[1]` 
tends to be the `r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[1]))$bus_route`, which takes `r round((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[1]))$avg_duration, 2)` 
minutes on average, and has an average crowded-ness of `r round((busiest_route %>% filter(bus_route == (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[1]))$bus_route, direction == least_crowded_overall$direction[1]))$avg_crowded, 2)`/5.

The quickest bus route from `r least_crowded_overall$direction[2]` tends to be the 
`r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[2]))$bus_route`, which takes `r round((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[2]))$avg_duration, 2)` 
minutes. This route typically has a `r round((busiest_route %>% filter(bus_route == (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[2]))$bus_route, direction == least_crowded_overall$direction[2]))$avg_crowded, 2)`/5 crowded-ness rating when I take it (i.e. 
in the collected data).

This suggests that if I am running late or need to reach my destination as soon as 
possible, I should consider taking these routes, and can anticipate how busy the 
bus will be to prepare myself to stand if necessary.


By using the data from my previous trips, I can observe trends in the duration and 
crowded-ness of various bus routes and transport types at various times of day. 
These insights enable me to plan ahead and make more informed decisions about what 
time I should leave to arrive at my destination at time, and the mode of transport 
I should use based on my values of efficiency and comfort.  

### Overall Insights:
* I can expect the `r avg_duration_by_mode$transport_mode[1]` 
to be quicker than the `r avg_duration_by_mode$transport_mode[2]` in general
* The `r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[2]))$bus_route` bus route is faster on average than the `r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == max(avg_duration), direction == least_crowded_overall$direction[2]))$bus_route` bus route by `r round(abs((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == max(avg_duration), direction == least_crowded_overall$direction[2]))$avg_duration), 2) - round(abs((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[2]))$avg_duration), 2)` minutes when travelling from `r least_crowded_overall$direction[2]`
* The `r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[1]))$bus_route` bus route is faster on average than the `r (avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == max(avg_duration), direction == least_crowded_overall$direction[1]))$bus_route` bus route by `r round(abs((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == max(avg_duration), direction == least_crowded_overall$direction[1]))$avg_duration), 2) - round(abs((avg_duration_by_route %>% filter(bus_route != "Car") %>% group_by(direction) %>% filter(avg_duration == min(avg_duration), direction == least_crowded_overall$direction[1]))$avg_duration), 2)` minutes when travelling from `r least_crowded_overall$direction[1]`
* I can expect my least crowded commute to be from  `r (busiest_route %>% filter(avg_crowded == min(avg_crowded)))$direction` on the `r (busiest_route %>% filter(avg_crowded == min(avg_crowded)))$bus_route`

